# Adrenergic Vasoactives {#sec-catechols}

Adrenergic drugs:

:::column-margin
This covers the pharmacology of specific catecholamines and sympathomimetics. The synthesis of endogenous catecholamines is covered under @sec-adrenal_hormones, whilst specifics of catecholamine receptor function is covered under @sec-adrenoreceptors.
:::

* Act via:
    * Dopamine receptors (D)
    * Adrenoreceptors (α and β)
* Can be:
    * **Direct-acting**\
    Stimulate the receptor.
    * **Indirect-acting**\
    Stimulate the release of noradrenaline to cause effects.
* Classified as either:
    * **Naturally-occurring catecholamines**
    * **Synthetic catecholamines**
    * **Synthetic sympathomimetics**\
    Drugs which act on adrenoreceptors but are not classified as catecholamines due to their chemical structure.


```{r}
source("scripts/drugtables.R")

DrugTable("adrenaline", "noradrenaline" , "dopamine", caption = "Naturally Occurring Adrenergic Vasoactives")
```



```{r}
DrugTable("dobutamine", "isoprenaline", "ephedrine", "metaraminol", "phenylephrine", caption = "Synthetic Adrenergic Vasoactives")
```


## Structure-Activity Relationships of Sympathomimetics

![](/resources/dopamine.svg){fig-alt="Chemical structure of dopamine, labelling the catechol ring, hydroxyl groups, ethylamine side chain, and alpha and beta carbons."}

Catecholamines consist of:

* A catechol ring\
A benzene ring with two hydroxyl groups in the 3 and 4 position.
    * Losing one hydroxyl group
        * Increases lipid solubility and decreases the potency 10-fold
        * Prevents metabolism by COMT, prolonging duration of action
    * Losing both hydroxyl groups decreases the potency 100-fold
	   * Changing the hydroxyl groups to the 3 and 5 position increases beta-2 selectivity when there is also a large substitution present on the amine group
* An ethylamine tail\
Consists of:
    * Beta carbon\
    The first carbon.
        * Adding a hydroxyl group decreases lipid solubility and CNS penetration
        * Adding any group increases alpha and beta selectivity
    * Alpha carbon\
    The second carbon.
        * Adding a group prevents metabolism by MAO, prolonging duration of action
        * Methylation increases indirect activity
    * Amine group\
    The terminal nitrogen.
        * Addition of a methyl group generally increases beta selectivity\
        As the chain length increases, so does the beta selectivity.


### Dopamine

![](/resources/dopamine.svg){fig-alt="Chemical structure of dopamine, labelling the catechol ring, hydroxyl groups, ethylamine side chain, and alpha and beta carbons."}

* Dopamine is the prototypical catecholamine, to which others are compared


### Noradrenaline


![](/resources/noradrenaline.svg){fig-alt="Chemical structure of noradrenaline, showing the hydroxyl group on the beta carbon in addition to the catechol hydroxyl groups."}

* Noradrenaline has a hydroxyl group added to the beta carbon, increasing its alpha selectivity


### Adrenaline

![](/resources/adrenaline.svg){fig-alt="Chemical structure of adrenaline, showing the catechol hydroxyl groups, beta-carbon hydroxyl group and methyl group on the amine."}

* Adrenaline is similar to noradrenaline with an additional hydroxyl group on the beta carbon
* Adrenaline also has a methyl group added to the terminal amine, increasing beta selectivity


### Metaraminol

![](/resources/metaraminol.svg){fig-alt="Chemical structure of metaraminol, showing one hydroxyl group on the phenol ring, a beta-carbon hydroxyl group and an alpha-carbon methyl group."}

* Metaraminol has an additional hydroxyl group on the beta carbon
* Metaraminol has only one hydroxyl group on the phenol ring, so:
  * It is no longer classified as a catecholamine
  * It is not metabolised by COMT, prolonging its duration of action
  * It has reduced potency, requiring administration in higher doses
* Metaraminol has an additional methyl group on the alpha carbon, preventing metabolism by MAO and further prolonging its duration of action


### Ephedrine

![](/resources/ephedrine.svg){fig-alt="Chemical structure of ephedrine, showing no hydroxyl groups on the benzene ring, a beta-carbon hydroxyl group and methyl groups on the alpha carbon and amine."}

* Like metaraminol, ephedrine has a hydroxyl group on the beta carbon and a methyl group on the alpha carbon
* Ephedrine has no hydroxyl groups on the phenol ring, further reducing its potency and increasing its elimination half-life
* Ephedrine has a methyl group on the amine, increasing its beta selectivity


***

## References

1. Brandis K. [The Physiology Viva: Questions & Answers](http://www.anaesthesiamcq.com/vivabook.php). 2003.
1. Chambers D, Huang C, Matthews G. Basic Physiology for Anaesthetists. Cambridge University Press. 2015.
1. Yartsev A. [Deranged Physiology - Structure of Synthetic Catecholamines](http://www.derangedphysiology.com/main/core-topics-intensive-care/critical-care-pharmacology/Chapter%203.1.3/structure-synthetic-catecholamines)
1. Peck TE, Hill SA. Pharmacology for Anaesthesia and Intensive Care. 4th Ed. Cambridge University Press. 2014.
