# Inhalational Anaesthetic Agents


:::dual
Describe the effects of inhalational agents on the cardiovascular, respiratory and central nervous systems
:::

:::dual
Describe the toxicity of inhalational agents
:::

:::dual
Describe the comparative pharmacology of ~~nitrous oxide~~, halothane, enflurane, isoflurane, desflurane, sevoflurane, xenon and ether
:::

*This section covers features and structures of inhalational anaesthetics. Structure-activity relationships are covered under [inhalational anaesthetics](inhalational_anaesthetic_agents.qmd#sa).*

: Common Features of Inhalational Anaesthetic Agents

+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| Property                             | Action                                                                                                    |
+======================================+===========================================================================================================+
| **Metabolism**                       | Hepatic CYP450 (CYP2E1) metabolises C-halogen bonds to release halogen ions (F^-^, Cl^-^, Br^-^), which   |
|                                      | can be nephrotoxic and hepatotoxic. The C-F bond is minimally metabolised compared to the C-Cl, C-Br, and |
|                                      | C-I bonds. All agents undergo hepatic oxidation, except for halothane which is reduced.                   |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **Resp**                             | All halogenated agents ↓ V~T~ and ↑ RR, with an overall ↓ in MV and therefore cause PaCO~2~ to ↑; and ↓   |
|                                      | sensitivity of central respiratory centres to CO~2~. Impairment of HPV may worsen V/Q matching and ↑      |
|                                      | shunt. ![Effect of volatile anaesthetics on the ventilatory response to                                   |
|                                      | hypoxia](/resources/volatile-po2-response.svg){width=55%}                                                 |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **CVS**                              | ↓ MAP (predominantly by ↓ in SVR due to NO release and Ca^2+^ channel blockade), ↓ inotropy due to Ca^2+^ |
|                                      | channel blockade.                                                                                         |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **CNS**                              | Hypnosis. ↓ CMRO~2~. Above 1 MAC there is **uncoupling** of the CBF-CMRO~2~ relationship, and CBF ↑       |
|                                      | despite ↓ CMRO~2~ due to cerebral vasodilation. ICP may mirror CBF changes.                               |
|                                      |                                                                                                           |
|                                      | All except halothane have some analgesic effect. ↓ EEG frequency such that θ- and δ-wave dominate the EEG |
|                                      | as depth ↑. May cause burst suppression. ![Effect of volatile anaesthetics on cerebral blood flow and     |
|                                      | metabolic rate](/resources/cbf-cmro2-volatile.svg){width=47%} ![Effect of sevoflurane concentration on    |
|                                      | cerebral blood flow and metabolic rate](/resources/cbf-cmro2-sevo.svg){width=47%}                         |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **MSK**                              | Muscle relaxation via blockade of Ca^2+^ channels. Additional augmentation of the effects of NMBD due to  |
|                                      | skeletal muscle vasodilation. May precipitate MH.                                                         |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **Renal**                            | Dose dependent ↓ in RBF, GFR, and UO secondary to ↓ in MAP and CO.                                        |
|                                      |                                                                                                           |
|                                      | Fluorinated ethers produce F^-^ ions when hepatically metabolised, which may produce high-output renal    |
|                                      | failure at serum concentrations **>50μmol/L**. This is probably only a concern with methoxyflurane (as it |
|                                      | has significant (>70%) hepatic metabolism) when used at anaesthetic doses.                                |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **GIT**                              | ↓ Hepatic blood flow.                                                                                     |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **GU**                               | Tocolysis.                                                                                                |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+
| **Toxic Effects**                    | Decreased fertility and increased risk of spontaneous abortion in operating theatre personnel.            |
+--------------------------------------+-----------------------------------------------------------------------------------------------------------+


## Comparison of Common Inhalational Agents

```{r}
source("scripts/drugtables.R")

DrugTable("sevoflurane", "isoflurane", "desflurane", caption = "Comparison of Common Inhalational Anaesthetics")
```

Sevoflurane:

![](/resources/sevoflurane-1.svg){fig-alt="Sevoflurane molecular structure." width=15pc}

Isoflurane:

![](/resources/isoflurane-2.svg){fig-alt="Isoflurane molecular structure." width=15pc}

Desflurane:

![](/resources/desflurane2.svg){fig-alt="Desflurane molecular structure." width=15pc}

![](/resources/desflurane-washout.svg){fig-alt="Time to a 90% decrease in sevoflurane or desflurane concentration plotted against duration of anaesthesia." width=100%}


## Comparison of Uncommon Inhalational Agents

```{r}
DrugTable("enflurane", "halothane", "xenon", caption = "Comparison of Uncommon Inhalational Anaesthetics")
```

Enflurane:

![](/resources/enflurane.svg){fig-alt="Enflurane molecular structure." width=15pc}

***

## References

1. Khan KS, Hayes I, Buggy DJ. [Pharmacology of anaesthetic agents II: inhalation anaesthetic agents](https://academic.oup.com/bjaed/article/14/3/106/340726/Pharmacology-of-anaesthetic-agents-II-inhalation). Continuing Education in Anaesthesia Critical Care & Pain, Volume 14, Issue 3, 1 June 2014, Pages 106–111.
2. Peck TE, Hill SA. Pharmacology for Anaesthesia and Intensive Care. 4th Ed. Cambridge University Press. 2014.\
3. Petkov V. Essential Pharmacology For The ANZCA Primary Examination. Vesselin Petkov. 2012.
4. Miller RD, Eriksson LI, Fleisher LA, Weiner-Kronish JP, Cohen NH, Young WL. Miller's Anaesthesia. 8th Ed (Revised). Elsevier Health Sciences.
Peck and Hill
5. Leslie RA, Johnson EK, Goodwin APL. Dr Podcast Scripts for the Primary FRCA. Cambridge University Press. 2011.
6. Law LS, Lo EA, Gan TJ. [Xenon Anesthesia: A Systematic Review and Meta-Analysis
of Randomized Controlled Trials](https://www.ncbi.nlm.nih.gov/pubmed/26273750). Anesth Analg. 2016 Mar;122(3):678-97.
