# Penicillins


* Penicillins are **bactericidal** antibiotics that prevent cell-wall synthesis by preventing cross-linking of peptidoglycans by replacing the natural substrate with their β-lactam ring
* Penicillins bind to penicillin binding proteins (**PBPs**) in the bacterial wall
* Penicillins only rarely achieve complete eradication of sensitive organisms without addition of a **synergistic antibiotic** (such as gentamicin)

: Common Pharmacological Features of Penicillins

|Property|Effect
|--|--|
|**Absorption**|Typically well absorbed orally. IM dosing tend to cause localised pain and irritation.|
|**Distribution**| Typically have good tissue penetration. Only cross the blood-brain barrier and enter bone if it is inflamed. Typically low protein binding (exception is flucloxacillin, which is 95% protein bound).|
|**Metabolism**| Typically small proportion is hepatically metabolised.
|**Elimination**| Majority (60-90%) is eliminated unchanged in urine predominantly by active tubular secretion, with renal clearance proportional to total renal plasma flow. A small quantity is secreted in bile.|

### Mechanisms of Resistance

* Alteration or protection of PBPs
  * Gram negative bacteria may have altered permeability of porins in their outer membrane, which protects the PDP
* Hydrolysis by **β-lactamase**-producing bacteria
  * **Clavulanic acid** and **tazobactam inhibit β-lactamase**, which can render otherwise resistant bacteria sensitive
  * Notably, flucloxacillin has a modified beta-lactam ring that is not sensitive to β-lactamases

: Comparison of Penicillin Classes

||Narrow spectrum, naturally occurring|Narrow spectrum, synthetic|Extended-spectrum|Antipseudomonal|
|--|--|--|--|--|
|**Examples**|**Benzylpenicillin**, phenoxymethylpenicillin|**Flucloxacillin**|**Ampicillin**, amoxacillin| **Piperacillin**, ticarcillin|
|**Indications**|Gram positives and anaerobes, particularly *streptococci* and meningococci. Also *listeria*, *Clostridia*, and *Treponema*.| Gram positive cocci, particularly *staphylococci* but also streptococci. | Gram positive, particularly *enterococci*. Some gram negative.|Gram positive, gram negative including pseudomonas.|
|**Other bits**| Highly bactericidal|Less active than benzylpenicillin on organisms sensitive to both.|Can penetrate some gram-negatives.|Gram negative cover.||

## Individual Penicillins

```{r}
source("scripts/drugtables.R")

DrugTable("benzylpenicillin", "phenoxymethylpenicillin", caption = "Naturally Occurring Penicillins")
```

```{r}
DrugTable("flucloxacillin", caption = "Anti-Staphylococcal Penicillins")
```

```{r}
DrugTable("ampicillin", "amoxicillin", caption = "Aminopenicillins")
```

```{r}
DrugTable("piperacillin", "ticarcillin", caption = "Antipseudomonal Penicillins")
```

***

## References

1. Peck TE, Hill SA. Pharmacology for Anaesthesia and Intensive Care. 4th Ed. Cambridge University Press. 2014.\
2. Barza M, Weinstein L. [Pharmacokinetics of the penicillins in man](https://www.ncbi.nlm.nih.gov/pubmed/797501). Clin Pharmacokinet. 1976;1(4):297-308.
3. Brunton LL, Chabner BA, Knollmann BC, eds. Goodman & Gilman's The Pharmacological Basis of Therapeutics. 12th Ed. McGraw-Hill. 2011.
4. CICM July/September 2007
