# Opioids


: Common Pharmacological Features of Opioids

|Property|Effect
|--|--|
|**Uses**|Analgesia, sedation, elimination of sympathetic response to laryngoscopy/surgical stress response|
|**Resp**|↓ CNS sensitivity to CO~2~ causing respiratory depression (↓ RR > ↓ V~T~) - ↑ reliance on hypoxic drive (therefore respiratory depression may be potentiated by high FiO~2~)|
|**CVS**|↓ HR. May ↓ BP due to histamine release (less with synthetic agents). ↑ PVR|
|**CNS**|Sedation, euphoria. Nausea and vomiting due to CTZ stimulation. Meiosis due to stimulation of the Edinger-Westphal nucleus. ↓ MAC up to 90%|
|**Renal**|↑ ADH, ↑ ureteric and sphincter tone
|**MSK**|Muscle rigidity, pruritus (especially with intrathecal administration)
|**Metabolic**|↓ ACTH, prolactin, gonadotrophic hormone secretion. ↑ ADH secretion|
|**GIT**|↓ Peristalsis and GIT secretions with subsequent constipation|
|**Immunological**|Impaired: chemotaxis, lymphocyte proliferation, and antibody production|

```{r}
source("scripts/drugtables.R")

DrugTable("morphine", caption = "Morphine")
```

```{r}
DrugTable("oxycodone", "buprenorphine", caption = "Comparison of Semisynthetic Opioids")
```


```{r}
DrugTable("fentanyl", "alfentanil", "remifentanil", caption = "Comparison of Synthetic Opioids")
```


***

## References

1. Peck TE, Hill SA. Pharmacology for Anaesthesia and Intensive Care. 4th Ed. Cambridge University Press. 2014.\
2. Smith S, Scarth E, Sasada M. Drugs in Anaesthesia and Intensive Care. 4th Ed. Oxford University Press. 2011.
3. Petkov V. Essential Pharmacology For The ANZCA Primary Examination. Vesselin Petkov. 2012.
4. ANZCA July/September 2010
5. Schug SA, Palmer GM, Scott DA, Halliwell R, Trinca J. [Acute Pain Management: Scientific Evidence](http://fpm.anzca.edu.au/documents/apmse4_2015_final). 4th Ed. 2015. Australian and New Zealand College of Anaesthetists and Faculty of Pain Medicine.
